---
title: "Sec_1_SA1-Percia-EDA"
author: "Kyte Daiter Percia"
date: "2025-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(gridExtra)
library(moments)
library(e1071)
library(MASS)
library(ggpmisc)
library(lm.beta)
library(lattice)
```

## Dataset: EDA_Ecommerce_Assessment.csv

```{r load-data}
df <- read.csv("C:/Users/john/Downloads/EDA_Ecommerce_Assessment.csv")
str(df)
summary(df)
```

---

## UNIT 1: Univariate Data Analysis

### 1. Distribution of Key Variables
We begin by visually examining the distribution of `Purchase_Amount`, `Number_of_Items`, and `Satisfaction_Score` using histograms and boxplots.

```{r hist-box}
par(mfrow = c(2,3))

hist(df$Purchase_Amount, main = "Histogram of Purchase Amount", col = "skyblue", xlab = "Purchase Amount", breaks = 30)
hist(df$Number_of_Items, main = "Histogram of Number of Items", col = "lightgreen", xlab = "Number of Items", breaks = 10)
hist(df$Satisfaction_Score, main = "Histogram of Satisfaction Score", col = "orange", xlab = "Satisfaction Score", breaks = 5)

boxplot(df$Purchase_Amount, main = "Boxplot of Purchase Amount", col = "skyblue", horizontal = TRUE)
boxplot(df$Number_of_Items, main = "Boxplot of Number of Items", col = "lightgreen", horizontal = TRUE)
boxplot(df$Satisfaction_Score, main = "Boxplot of Satisfaction Score", col = "orange", horizontal = TRUE)

par(mfrow = c(1,1))
```

**Interpretation:** All three variables exhibit right-skewed distributions. Extreme values exist, particularly in `Purchase_Amount`. `Satisfaction_Score` appears more categorical.

---

### 2. Summary Statistics for Purchase Amount
We compute central tendency and spread measures for `Purchase_Amount`.

```{r summary-stats}
mean_purchase <- mean(df$Purchase_Amount, na.rm = TRUE)
median_purchase <- median(df$Purchase_Amount, na.rm = TRUE)
mode_purchase <- as.numeric(names(sort(table(df$Purchase_Amount), decreasing = TRUE)[1]))
var_purchase <- var(df$Purchase_Amount, na.rm = TRUE)
sd_purchase <- sd(df$Purchase_Amount, na.rm = TRUE)
iqr_purchase <- IQR(df$Purchase_Amount, na.rm = TRUE)

cat("Mean:", mean_purchase, "\n")
cat("Median:", median_purchase, "\n")
cat("Mode:", mode_purchase, "\n")
cat("Variance:", var_purchase, "\n")
cat("Standard Deviation:", sd_purchase, "\n")
cat("IQR:", iqr_purchase, "\n")
```

**Interpretation:** The mean and median are close, indicating a moderately symmetric distribution. High variance and standard deviation highlight variability.

---

### 3. Density Plots by Gender
We explore how `Browsing_Time` and `Purchase_Amount` differ across genders.

```{r gender-density}
plot1 <- ggplot(df, aes(x = Browsing_Time, fill = Gender)) +
  geom_density(alpha = 0.5) + ggtitle("Browsing Time by Gender")

plot2 <- ggplot(df, aes(x = Purchase_Amount, fill = Gender)) +
  geom_density(alpha = 0.5) + ggtitle("Purchase Amount by Gender")

grid.arrange(plot1, plot2, ncol = 2)
```

**Interpretation:** Minor differences are observed between genders. Males show a slightly higher peak in spending.

---

### 4. Skewness and Transformation
We evaluate the impact of logarithmic and square root transformations on `Browsing_Time`.

```{r transform-skewness}
df <- df %>% mutate(log_Browsing_Time = log1p(Browsing_Time), sqrt_Browsing_Time = sqrt(Browsing_Time))

original_skew <- skewness(df$Browsing_Time)
log_skew <- skewness(df$log_Browsing_Time)
sqrt_skew <- skewness(df$sqrt_Browsing_Time)

cat("Original Skewness:", original_skew, "\n")
cat("Log-transformed Skewness:", log_skew, "\n")
cat("Square Root Skewness:", sqrt_skew, "\n")
```

**Interpretation:** Original `Browsing_Time` is nearly symmetric. Log transformation introduces left skew.

---

### 5. Simple Linear Regression
We model `Purchase_Amount` using `Browsing_Time` as a predictor.

```{r simple-regression}
model <- lm(Purchase_Amount ~ Browsing_Time, data = df)
summary(model)
```

**Interpretation:** Coefficient for `Browsing_Time` is not statistically significant. R-squared is near zero, indicating a poor fit.

---

### 6. Regression Visualization

```{r scatter-lm}
ggplot(df, aes(x = Browsing_Time, y = Purchase_Amount)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  ggtitle("Purchase Amount vs Browsing Time")
```

**Interpretation:** Scattered points around a flat line confirm weak or no linear relationship.

---

## UNIT 2: Bivariate Data Analysis

### 7. Scatter Plot: Purchase Amount vs. Number of Items

```{r items-vs-amount}
ggplot(df, aes(x = Number_of_Items, y = Purchase_Amount)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  ggtitle("Purchase Amount vs Number of Items")
```

**Interpretation:** A positive trend is visible—more items generally lead to higher purchase amounts.

---

### 8. Polynomial Regression Comparison

```{r poly-regression}
poly_model <- lm(Purchase_Amount ~ poly(Browsing_Time, 2, raw = TRUE), data = df)
summary(poly_model)
```

**Interpretation:** The polynomial model fits better than a simple linear one, showing a peak spending time.

---

### 9. LOESS Smoothing

```{r loess}
ggplot(df, aes(x = Browsing_Time, y = Purchase_Amount)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", color = "purple") +
  ggtitle("LOESS: Purchase Amount vs Browsing Time")
```

**Interpretation:** LOESS curve captures subtle nonlinear patterns more effectively.

---

### 10. Robust Regression

```{r robust}
huber_model <- rlm(Purchase_Amount ~ Browsing_Time, data = df)
summary(huber_model)
```

**Interpretation:** Robust regression reduces the impact of outliers, offering more stable results than OLS.

---

### 11. Interaction Plot: Category x Browsing Time

```{r interaction}
ggplot(df, aes(x = Browsing_Time, y = Purchase_Amount, color = Category)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("Interaction: Browsing Time by Category")
```

**Interpretation:** Different product categories influence spending behavior across browsing times.

---

### 12. Coplots for Categories

```{r coplots}
coplot(Purchase_Amount ~ Browsing_Time | Category, data = df, panel = panel.smooth)
```

**Interpretation:** Some categories like electronics show strong associations with longer browsing.

---

### 13. Multiple Regression Analysis

```{r multiple-regression}
multi_model <- lm(Purchase_Amount ~ Browsing_Time + Number_of_Items + Satisfaction_Score, data = df)
summary(multi_model)
lm.beta(multi_model)
```

**Interpretation:** `Number_of_Items` and `Satisfaction_Score` are strong predictors. `Browsing_Time` remains weak.

---

### 14. Model Selection via Stepwise Regression

```{r stepwise}
stepwise_model <- step(multi_model, direction = "both")
summary(stepwise_model)
```

**Interpretation:** Stepwise selection confirms variable importance. Final model prioritizes `Number_of_Items` and `Satisfaction_Score`.
