---
title: "Sec1_FA4-Percia-EDA"
author: "Kyte Daiter Percia"
date: "2025-05-13"
output: html_document
---

```{r setup}
mortality_data <- read.csv("C:/Users/john/Downloads/mortality_by_latitude.csv")

library(ggplot2)
```

### Formative Assessment 4

Using the dataset mortality by latitude and diamonds data set, we will be ble to compare two plot residuals in the two models to see which is more fit.

#### Number 1: Mortality by Latitude Data

```{r one_1}
ggplot(mortality_data, aes(x = temperature, y = mortality_index)) +
  geom_point(size = 3, color = "darkred") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linetype = "dashed") +
  labs(title = "Mortality Index vs. Mean Average Temperature",
       x = "Mean Average Temperature (F)",
       y = "Mortality Index") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The relationship is **hollow-up** (concave upward) as mortality increases with temperature. This suggests an exponential relationship rather than logarithmic.

**First try** linearizing the data:

```{r one_2}
# Square temperature
mortality_data$temp_squared <- mortality_data$temperature^2

# Plot transformed relationship
ggplot(mortality_data, aes(x = temp_squared, y = mortality_index)) +
  geom_point(size = 3, color = "darkgreen") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "orange") +
  labs(
    title = "Mortality Index vs. Squared Temperature",
    x = "Temperature Squared (°F²)",
    y = "Mortality Index"
  ) +
  theme_minimal()
```

Seems like squaring is still not enough.

**Second try** linearizing the data:

```{r one_3}
# Apply exponential scaling (e.g., exp(temperature/10) to avoid extreme values)
mortality_data$temp_exp <- exp(mortality_data$temperature / 10)

ggplot(mortality_data, aes(x = temp_exp, y = mortality_index)) +
  geom_point(size = 3, color = "darkgreen") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "orange") +
  labs(
    title = "Mortality Index vs. Exponential Temperature",
    x = "exp(Temperature / 10)",
    y = "Mortality Index"
  ) +
  theme_minimal()
```

Now, to verify if residuals are patternless.

```{r one_4}
# Fit model (using squared temperature)
model <- lm(mortality_index ~ temp_squared, data = mortality_data)

# Residual plot
ggplot(mortality_data, aes(x = fitted(model), y = resid(model))) +
  geom_point(size = 3, color = "purple") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  labs(
    title = "Residuals vs. Fitted Values (Squared Temp)",
    x = "Fitted Values",
    y = "Residuals"
  ) +
  theme_minimal()
```

The residual plot shows a **random scatter** meaning the transformation worked.

#### Number 2: Diamonds Dataset

```{r two_1}
data(diamonds)

## Load Data & Apply Log Transformation

# Subset for faster computation (e.g., 10% of data)
set.seed(123)
diamonds_sub <- diamonds[sample(nrow(diamonds), size = 0.1 * nrow(diamonds)), ]

# Add log(price)
diamonds_sub$log_price <- log(diamonds_sub$price)

##  Plot with LOESS Smoother
# Default parameters
ggplot(diamonds_sub, aes(x = carat, y = log_price)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "loess", span = 0.75, degree = 2, 
              color = "red", se = FALSE, linewidth = 1) +
  labs(title = "Log(Price) vs. Carat (Default LOESS: span=0.75, degree=2)",
       x = "Carat", y = "Log(Price)") +
  theme_minimal()
```

Trying other parameters.

```{r two_2}
#Tuning span
# Example: Larger span (smoother)
ggplot(diamonds_sub, aes(x = carat, y = log_price)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", span = 1.0, color = "darkgreen", se = FALSE) +
  labs(title = "Span = 1.0 (Smoother)")

#Tuning degree
# Example: Degree = 1 (linear local fits)
ggplot(diamonds_sub, aes(x = carat, y = log_price)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", span = 0.75, degree = 1, 
              color = "purple", se = FALSE) +
  labs(title = "Degree = 1 (Linear Local Fits)")

# Final recommended plot
ggplot(diamonds_sub, aes(x = carat, y = log_price)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "loess", span = 0.5, degree = 2,
              color = "red", se = FALSE, linewidth = 1) +
  labs(title = "Recommended: Log(Price) vs. Carat (span=0.5, degree=2)",
       x = "Carat", y = "Log(Price)") +
  theme_minimal()
```

After testing, a **span = 0.5** and **degree = 2** works well for this data as it balances capturing the upward curve in log(price) at higher carats without overfitting noise and the relationship (log(price) rises faster with carat), so quadratic local fits adapt better than linear.

#### Number 3: Comparing plot of the residuals in the two models

Comparing them starting from sratch, using the diamonds dataset of R.

```{r three_1}
library(ggplot2)
library(patchwork)  # For combining plots
data(diamonds)

# Subset for speed (10% of data)
set.seed(123)
diamonds_sub <- diamonds[sample(nrow(diamonds), 1000), ]
diamonds_sub$log_price <- log(diamonds_sub$price)  # Log-transform price

# Model 1: LOESS smoother (span=0.5, degree=2)
loess_fit <- loess(log_price ~ carat, data = diamonds_sub, span = 0.5, degree = 2)
diamonds_sub$loess_resid <- residuals(loess_fit)  # Add LOESS residuals

# Model 2: Polynomial (cubic) + step function
diamonds_sub$carat_bin <- cut(diamonds_sub$carat, breaks = c(0, 1, 2, 3, 5))  # Binning
poly_step_fit <- lm(log_price ~ poly(carat, 3) + carat_bin, data = diamonds_sub)
diamonds_sub$poly_step_resid <- residuals(poly_step_fit)  # Add polynomial+step residuals

# LOESS residuals plot
p1 <- ggplot(diamonds_sub, aes(x = carat, y = loess_resid)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "LOESS Residuals", x = "Carat", y = "Residuals") +
  theme_minimal()

# Polynomial + Step residuals plot
p2 <- ggplot(diamonds_sub, aes(x = carat, y = poly_step_resid)) +
  geom_point(alpha = 0.5, color = "green") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Polynomial + Step Residuals", x = "Carat", y = "Residuals") +
  theme_minimal()

# Combine plots
p1 + p2 + plot_layout(ncol = 2)
```

Interpretation:

- **LOESS performs better** as residuals are randomly distributed, indicating it captures the relationship well. Larger residuals are acceptable for extreme carat values (rare in the dataset).
- There are **issuess** on **Polynomial + Step** as residuals shows a **U-shape**, meaning it systematically underfits wherein it underpredicts log(price) for very small/large carats or discontinuities at bin edges.
- Two actionable insights is first, use LOESS for pricing models where smotth trends matter and also at the same time avoid polynomial + step unless bins align with business rules like fixed price tiers.


